---
title: "Assignment 2"
author: "Ignacio Amaya"
date: "19 de diciembre de 2015"
output: pdf_document
---

DATA DESCRIPTION:
Monthly interest rate of unsubsidised (no subvencionadas) loans for home purchase (percentage).
Monthly time series from january 2000 to december 2004.
Source: Banco de España (www.bde.es)
```{r}
library(TSA)
library(forecast)
library(astsa)
dataAsigment2 <- read.csv("~/MIS DOCUMENTOS/DATA SCIENCE MASTER (EIT DIGITAL)/Intelligent Data Analysis (statistics)/Assignment 2 (time series)/dataAsigment2.csv", sep=";")
```

```{r}
g12data=ts(dataAsigment2$Tipo,start=c(1995,1), end=c(2013,9),frequency=12)
str(g12data)
plot(g12data)
tsdisplay(g12data, plot.type="partial")
tsdisplay(g12data, plot.type="scatter") #The one Arminda likes most
```

Let's apply BoxCox transformations to check if it is possible to apply a logaritmic transformation to our time series.
```{r}
BoxCox.lambda(g12data,lower=0, upper=2)
tsdisplay(log(g12data), plot.type="scatter")
```
As it is not close to zero we can't apply this transformation. There is a slight improvement because we can see in the third plot that the values are more spread across the plot.

We can observe that we have a linear trend (decreasing values over time) and a seasonal component as after some years decreasing the time series goes up again (and that happens several times).

```{r}
## Holt-Winters smoothing with series with linear trend and seasonal variation

smoothing1=HoltWinters(g12data,gamma=FALSE,beta=FALSE)
smoothing2=HoltWinters(g12data,gamma=FALSE)
smoothing3=HoltWinters(g12data)
smoothing4=HoltWinters(g12data,seasonal = "mult")
  
forecast1=predict(smoothing1, n.ahead=6, prediction.interval=TRUE, level=0.95)
forecast2=predict(smoothing2, n.ahead=6, prediction.interval=TRUE, level=0.95)
forecast3=predict(smoothing3, n.ahead=6, prediction.interval=TRUE, level=0.95)
forecast4=predict(smoothing4, n.ahead=6, prediction.interval=TRUE,level=0.95)

plot(smoothing1, forecast1)
plot(smoothing2, forecast2)
plot(smoothing3, forecast3)
plot(smoothing4, forecast4)
```

Let's try to decompose this time series.
```{r}
decomp=decompose(g12data)
plot(decomp)
decompLog=decompose(log(g12data))
plot(decompLog)
### Decomposition of time series, multiplicative decomposition
decompMult=decompose(g12data,type="multiplicative")
plot(decompMult)
```

```{r}
tsdisplay(decomp$random,plot.type="scatter")
tsdisplay(decompLog$random,plot.type="scatter")
tsdisplay(decompMult$random,plot.type="scatter")
```

```{r}
stl=stl(g12data, s.window="periodic", robust=TRUE)
plot(stl)
par(mfrow=c(3,1))
fcst1=forecast(stl, method="naive")
plot(fcst1)
fcst2=forecast(stl, method="arima")
plot(fcst2)
fcst3=forecast(stl, method="rwdrift")
plot(fcst3)
fcst3$mean
```

```{r}
par(mfrow=c(2,1))
## in the logarithmic scale (log(a.ts))
lstl=stl(log(g12data), s.window="periodic", robust=TRUE)
fcst4=forecast(lstl, method="arima")
plot(fcst4)
## in the original scale 
fcst5=stlf(g12data,method="arima",lambda=BoxCox.lambda(g12data))
plot(fcst5)
```

No parece que haya una mejora apreciable al realizar la transformación logarítmica.
```{r}
par(mfrow=c(2,1))
stl_remainder=stl$time.series[,3]
lstl_remainder=lstl$time.series[,3]
plot(stl_remainder)
plot(lstl_remainder)
acf2(stl_remainder)
acf2(lstl_remainder)
```

There's not a clear seasonality pattern in the season plot.
But in each season trough the years (monthplot) some patterns can be found
```{r}
par(mfrow=c(1,1))
monthplot(g12data)
seasonplot(g12data)
```


It may help us in determining the period of the series, if any.
No idea!!!

```{r}
mvspec(g12data)
```

Exploring the autocorrelation structure of time series

```{r}
par(mfrow=c(3,1))
plot(g12data)
acf(g12data)
pacf(g12data)
```

```{r}
plot(g12data)
plot(diff(g12data,12))
plot(diff(diff(g12data,12))) #it looks more random
plot(diff(diff(diff(g12data,12)))) #it looks more random (seems the best one)

# exploring ggb
tsdisplay(g12data, plot.type="scatter")
tsdisplay(diff(g12data), plot.type="scatter") 
tsdisplay(diff(g12data,12), plot.type="scatter") 
tsdisplay(diff(diff(g12data),12), plot.type="scatter") 
tsdisplay(diff(diff(g12data)), plot.type="scatter") 
tsdisplay(diff(diff(diff(g12data),12)), plot.type="scatter")
tsdisplay(diff(diff(diff(g12data))), plot.type="scatter")
tsdisplay(diff(diff(diff(diff(g12data),12))), plot.type="scatter") #not improved
tsdisplay(diff(diff(diff(diff(g12data)))), plot.type="scatter") #not improved (bad)
```

Check if it's stationary to see when we can stop differentiating
Can only be applied to seasonality adjusted data (in our case we didn't find a 
seasonality pattern so it's ok)
```{r}
adf.test(diff(g12data,12))
adf.test(diff(g12data))
adf.test(diff(diff(g12data,12)))
adf.test(diff(diff(g12data)))

```

```{r}
#Check when to stop using standard deviation
sd(g12data)
sd(diff(g12data)) #BETTER
sd(diff(g12data,12))
sd(diff(diff(g12data,12))) 
sd(diff(diff(g12data))) #Smallest!!! Not seasonality applied
sd(diff(diff(diff(g12data,12))))
sd(diff(diff(diff(g12data))))

```

```{r}
tsdisplay(diff(g12data), plot.type="scatter") 
tsdisplay(diff(g12data), plot.type="partial") 
tsdisplay(diff(diff(g12data)), plot.type="scatter") 
tsdisplay(diff(diff(g12data)), plot.type="partial") 
tsdisplay(diff(diff(log(g12data))), plot.type="scatter")  #Improves slightly (but model is more complicated so we decide not to take logs)
```

ARIMA(p,d,q)(P,D,Q)
p->past observations
q->past errors
d->diferentiation order

D-> seasonal differentiation order (D=0)
```{r}
#We start from 2 normal diffs and none in the seasonal part because
#any diff with period was done before to improve the sd
g12arima.1=Arima(g12data,order=c(0,2,0),include.mean=1,seasonal=list(order=c(0,0,0)))

#The best one is Arima3
g12arima.2=Arima(g12data,order=c(1,2,0),include.mean=1,seasonal=list(order=c(0,0,0)))
g12arima.3=Arima(g12data,order=c(1,2,1),include.mean=1,seasonal=list(order=c(0,0,0))) 
g12arima.4=Arima(g12data,order=c(0,2,1),include.mean=1,seasonal=list(order=c(0,0,0)))

g12arima.4=Arima(g12data,order=c(1,0,1),include.mean=1,seasonal=list(order=c(0,1,0)))

#The best one now is Arima7
g12arima.5=Arima(g12data,order=c(1,2,1),include.mean=1,seasonal=list(order=c(1,0,1))) 
g12arima.6=Arima(g12data,order=c(2,1,1),include.mean=1,seasonal=list(order=c(1,0,1))) 
g12arima.7=Arima(g12data,order=c(2,1,0),include.mean=1,seasonal=list(order=c(1,0,1))) 

#my-automatic (d=1 D=0): 1 0 1 0 (sarima)
sarima(g12data,1,1,0,1,0,0,12)
#my-automatic (d=2 D=0): 1 1 1 1 (sarima) -> more complex and less aic
sarima(g12data,1,2,1,1,0,1,12)

#The one obtained with my-automatic is better (less aic and less complex)
g12arima.8=Arima(g12data,order=c(1,1,0),include.mean=1,seasonal=list(order=c(1,0,0))) 
auto.arima(g12data)

AIC(g12arima.1)
AIC(g12arima.2)
AIC(g12arima.3) 
AIC(g12arima.4)
AIC(g12arima.5)
AIC(g12arima.6)
AIC(g12arima.7) #This is the best one (it is the lowest one)
AIC(g12arima.8) #This is the best one (it is the lowest one being simpler)

```

We'll keep the last one

Maybe compare them with bad ones?? WTF?
```{r}
tsdiag(g12arima.8)
acf2(g12arima.8$residuals)

```

```{r}
#correlations between model coefficients
cov2cor(g12arima.8$var.coef) #Not greater than 0.8 so it is good
#testing independence of the residuals (apparently they are not independent???)
#CHECK THE P-VALUES (are them white noise or not???)
LB.test(g12arima.8) #it seems ok because in the bubble example the values are the same
#or 
LB.test(g12arima.8, lag=20)
#normality of the residuals
jarque.bera.test(residuals(g12arima.8)) #less than 0.05, they are not normal (seems bad)
#Checks the independence in a time series (greater than 0.05 seems alright)
Box.test(residuals(g12arima.8),lag=12)

#So the residuals seem normal but nott independent

## Forecasting with the model lynx.3
predict(g12arima.8,n.ahead=20)
TSA::plot.Arima(g12arima.8) #TSA package
plot(forecast(g12arima.8,h=12))
forecast(g12arima.8,h=12) #same predictions plotted with the TSA package
```
We could try logs to see if the bad distribution of the residuals is fixed.

```{r}
g12arima.9=Arima(log(g12data),order=c(1,1,0),include.mean=1,seasonal=list(order=c(1,0,0))) 
AIC(g12arima.9) #aic lower than before (by a lot)

cov2cor(g12arima.9$var.coef) #Not greater than 0.8 so it is good
#testing independence of the residuals (apparently they are not independent???)
#CHECK THE P-VALUES (are them white noise or not???)
LB.test(g12arima.9) #it seems ok because in the first example the values are the same (it is lower now than before but it doesn't seem important if harque.bera is good now)
#or 
LB.test(g12arima.9, lag=20)
#normality of the residuals
jarque.bera.test(residuals(g12arima.9)) #less than 0.05, they are not normal (seems bad)
#Checks the independence in a time series (greater than 0.05 seems alright)
Box.test(residuals(g12arima.9),lag=12) #She uses it instead of jarque.vera when the other doesn't work (as this works we are ok)
```

We don't take logs as it doesn't improve anything
